Single Bivalent Level Analysis

### 2020-01-24 doc/Bivalent_Analysis.Rmd
spell checks

Bivalent level traits and metrics have been added in the src/Setup_BivData.Rmd script. These observations are from the automated image analysis algorithm and have been currated (threw out incorrect algorithm output).

@@

- Look into ordering the mouse strains to highlight the MSM and PWD differences, renameing the chrm type labels, spacing out the x axis labels or removing some of the other mice.

@@

In an attempt to test for and metigate noise based on chromosome identity/physical size, I tried to isolate bivalents which are in the top quartille for SC length from their cells.

Since the majority of cells from the Automated BivData set do not have the complete completement of bivalents this subset is much smaller and might have 'smaller' bivalents from cells missing the longer bivalents. But assuming the algorithm's identification of bivalents in random, with a large enough sample size, hopefully this error would be minimized.

@@

Each point is a bivalent ploted by cell on the x axis. X's are the 4th quartile, big point is the mean and smaller black point is the median. I'm using these to compare the patterns of these statistics in the automated dataset which is missing some bivalent data.

@@

All the plots above show the distrtributions of manually whole cell measured SC lengths compared to the SC length distributions from the automated bivalent data. It shows the amount of within cell variance across strains. There is a bit of variance across the SC length distributions in the PWD females.

I'm purposing to take all the automated biv measures above the cell's 4th quartille as a dataset for 'longest' bivalents which we would predict to exclude the shortest chromsomes which would have stronger chromsome size effects.

But this dataset might be noisy, given the amount of variance in the SC length distributions across cells (PWD females, WSB females). But it's easier to just run another set of analysis.

@@

The DF **real.long.bivData** contains `r length(real.long.bivData$Obj.ID)` bivalent measures. full dataset is 
`r length(Curated_BivData$Obj.ID)`. This is the breakdown of bivalent observations by category `r data.frame(table(droplevels(real.long.bivData$category)))`.

@@

These plots show the SC lengths for the 'long SC data set'. They are supposed to be the longest 4-5 SC from cells where I could get good measures. These longer bivalents are useful because their patterns shouldn't be affect by chromsome size effect (which effects, CO position). Hopefully this data set will have less noise from chromosome identity, but there was still data missing (they don't come from whole cell measures).

@@

- Of all female single bivalents observations, 5% are XXs (1 of 20).

@@

- The average % of XX for whole cell SC (sum(all bivalents)) can be calculated from the whole.cell dataset. Lets guess 12% of a cell's total SC area is XX.

@@

- The data are continous and closer to normal/gaussian expectations than the MLH1 counts.

@@

- Limited work to add to the dataset if I can justify that the collection of single bivalents are random and equivilent across mice.

@@

1) SC length will be sexually dimorphic (sex effect will be signicant)(cite Lynn)

@@

@@

TRY ADDING NESTED boxplots for hand.foci.count

@@

remeber the long biv data set in **real.long.bivData**

@@

- Sex is a significant effect for SC length
The results seem to indicate that sex is a significant factor. Consider writing a subsampling approach (randomize / permute a data set of BivData)

@@

Preidction, High rec males have longer SC.

@@

In a logistic regression, mouse average SC length is slightly predictive telling if a mouse is in a high or low Rec strain.  I counldn't get the Mixed models working for the male polymorphism predictions...

@@

These seond set of models are made using the lmer()

@@

Random effect of strain is not significnt for ABS IFD, and only slightly significant for the IFD.PER

@@

The random strain effect was tested for raw and normalized IFD, neither are signifcant.

@@

None of the logistic regression models for ABS or PER IFD lengths are sigificant, even when just the Musc strains are used.

@@

Preliminary results from an independant dataset indecated that PWD had longer IFDs, which goes agaisnt the simple prediction of more COs ~ denser spacing of foci on the same bivalent. This also indecated that interference distance may evolved in the house mouse complex.

@@

These bosplots show that females have a much more medial poition of single foci bivalents, (much closer to 50% compared to males). They also show that Musc males' Foci1 position is slightly more central / medial compared to the same type of positions in the Dom male strains. MOLF males have much more medial positions than other strains.

@@

The logic of how the sis-co-ten metric is outlined in the figure below. The goal is to use this metric to model different tension ative cohesin amounts as a consequence of different numbers and placements of chiasmata/CO. This metric is calculated using SC area as a proxy to the amount of cohesion at metaphase.

@@

Males have much clearer seperation of siscoten across chrm classes. This is emphasized when SC length is also plotted. It seems like musc males have higher amounts of this metric compared to Dom males (which might)

To formally test the differences in sis-co-ten I plan to write a subsampling / permutation loop to compare the mean(sis.co.ten) of the same numbers of bivalents of the same class.

@@

Yes, model od predicting the mouse average siscoten, sex and sex-subp interaction are significant factors. The Random strain effect is also significant.

@@

My metric for telomere and centromere distance measure the distance of the nearest foci to the ends of the bivalent (SC). In the plots below each point is a single bivalent. I choose not to use the mark for centromere because it seems noisy and inconsistant...

@@

Dr. Broman suggested that the Smirnov K /(curve comparison) wasn't the best test to differences in general CO position. He sugggested doing simple t-test for the positions

>>>>

Bivalent level traits and metrics have been added in the src/Setup_BivData.Rmd script. These observations are from the automated image analysis algorithm and have been curated (threw out incorrect algorithm output).

@@

- Look into ordering the mouse strains to highlight the MSM and PWD differences, renaming the chrm type labels, spacing out the x axis labels or removing some of the other mice.

@@

In an attempt to test for and mitigate noise based on chromosome identity/physical size, I tried to isolate bivalents which are in the top quartille for SC length from their cells.

Since the majority of cells from the Automated BivData set do not have the complete complement of bivalents this subset is much smaller and might have 'smaller' bivalents from cells missing the longer bivalents. But assuming the algorithm's identification of bivalents in random, with a large enough sample size, hopefully this error would be minimized.

@@

Each point is a bivalent plotted by cell on the x axis. X's are the 4th quartile, big point is the mean and smaller black point is the median. I'm using these to compare the patterns of these statistics in the automated data set which is missing some bivalent data.

@@

All the plots above show the distributions of manually whole cell measured SC lengths compared to the SC length distributions from the automated bivalent data. It shows the amount of within cell variance across strains. There is a bit of variance across the SC length distributions in the PWD females.

I'm purposing to take all the automated biv measures above the cell's 4th quartille as a data set for 'longest' bivalents which we would predict to exclude the shortest chromsomes which would have stronger chromosome size effects.

But this data set might be noisy, given the amount of variance in the SC length distributions across cells (PWD females, WSB females). But it's easier to just run another set of analysis.

@@

The DF **real.long.bivData** contains `r length(real.long.bivData$Obj.ID)` bivalent measures. full data set is 
`r length(Curated_BivData$Obj.ID)`. This is the breakdown of bivalent observations by category `r data.frame(table(droplevels(real.long.bivData$category)))`.

@@

These plots show the SC lengths for the 'long SC data set'. They are supposed to be the longest 4-5 SC from cells where I could get good measures. These longer bivalents are useful because their patterns shouldn't be affect by chromosome size effect (which effects, CO position). Hopefully this data set will have less noise from chromosome identity, but there was still data missing (they don't come from whole cell measures).

@@

- Of all female single bivalents observations, 5% are XX (1 of 20).

@@

- The average % of XX for whole cell SC (sum(all bivalents)) can be calculated from the whole.cell data set. Lets guess 12% of a cell's total SC area is XX.

@@

- The data are continuous and closer to normal/Gaussian expectations than the MLH1 counts.

@@

- Limited work to add to the data set if I can justify that the collection of single bivalents are random and equivalent across mice.

@@

1) SC length will be sexually dimorphic (sex effect will be significant)(cite Lynn)

@@

  B) **sis**ter **co**hesion **ten**sion (sis-co-ten) will be sexually dimorphic as it reflects the general property of uniform vs telomere/biased CO positioning.

@@

TRY ADDING NESTED box plots for hand.foci.count

@@

remember the long biv data set in **real.long.bivData**

@@

- Sex is a significant effect for SC length
The results seem to indicate that sex is a significant factor. Consider writing a sub sampling approach (randomize / permute a data set of BivData)

@@

Prediction, High rec males have longer SC.

@@

In a logistic regression, mouse average SC length is slightly predictive telling if a mouse is in a high or low Rec strain.  I couldn't get the Mixed models working for the male polymorphism predictions...

@@

These second set of models are made using the lmer()

@@

Random effect of strain is not significant for ABS IFD, and only slightly significant for the IFD.PER

@@

The random strain effect was tested for raw and normalized IFD, neither are significant.

@@

None of the logistic regression models for ABS or PER IFD lengths are significant, even when just the Musc strains are used.

@@

Preliminary results from an independent data set indicated that PWD had longer IFDs, which goes against the simple prediction of more COs ~ denser spacing of foci on the same bivalent. This also indicated that interference distance may evolved in the house mouse complex.

@@

These box plot show that females have a much more medial position of single foci bivalents, (much closer to 50% compared to males). They also show that Musc males' Foci1 position is slightly more central / medial compared to the same type of positions in the Dom male strains. MOLF males have much more medial positions than other strains.

@@

The logic of how the sis-co-ten metric is outlined in the figure below. The goal is to use this metric to model different tension active cohesion amounts as a consequence of different numbers and placements of chiasmata/CO. This metric is calculated using SC area as a proxy to the amount of cohesion at metaphase.

@@

Males have much clearer separation of siscoten across chrm classes. This is emphasized when SC length is also plotted. It seems like musc males have higher amounts of this metric compared to Dom males (which might)

To formally test the differences in sis-co-ten I plan to write a sub sampling / permutation loop to compare the mean(sis.co.ten) of the same numbers of bivalents of the same class.

@@

Yes, model  predicting the mouse average siscoten, sex and sex-subp interaction are significant factors. The Random strain effect is also significant.

@@

My metric for telomere and centromere distance measure the distance of the nearest foci to the ends of the bivalent (SC). In the plots below each point is a single bivalent. I choose not to use the mark for centromere because it seems noisy and inconsistent...

@@

Dr. Broman suggested that the Smirnov K /(curve comparison) wasn't the best test to differences in general CO position. He suggested doing simple t-test for the positions
